---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Open Wave Height Logger data processing with R

The package `owhlR` contains functions to post-process raw csv data files produced by an Open Wave Height Logger (OWHL) pressure sensor data logger. See the vignette `OWHL-file-handling-vignette` for a walk-through of the steps.

This package depends on the `oceanwaves` package 
[https://github.com/millerlp/oceanwaves](https://github.com/millerlp/oceanwaves) and its 
downstream dependencies for the calculation of wave statistics after the raw pressure data have been imported and quality-checked. 

To install the development version of this package from within R, first install the package `devtools`
[https://CRAN.R-project.org/package=devtools](https://CRAN.R-project.org/package=devtools)
and then install this package from Github:

```{r install_github, eval = FALSE}
install.packages('devtools')
library(devtools)
install_github('millerlp/owhlR')
```

For more information on the Open Wave Height Logger project, 
see [https://lukemiller.org/?s=owhl](https://lukemiller.org/?s=owhl) and the Github repository: [https://github.com/millerlp/OWHL](https://github.com/millerlp/OWHL)
